\documentclass[a4paper,12pt]{article}
\parindent 0pt
\parskip 1mm
\usepackage{amsmath}
\usepackage[dvips]{epsfig}

\begin{document}

\begin{center}

{\Large\bf MA 555 - Numerical Analysis}

\bigskip

{\large\bf Assignment \# 6}
\smallskip

{\large\bf John Joseph}
\end{center}

{\bf Problem 1}
\smallskip

We aim to numerically solve the IVP $yy'=1$, $y(0)=1$ using three methods. 
\smallskip

Euler Explicit

\begin{eqnarray*}
yy'=1 \\
y'=\frac{1}{y} \\
y_1=y_0+(\frac{1}{y_0}) \tau=1+\tau
\end{eqnarray*}

Euler Implicit

\begin{eqnarray*}
y_1=y_0+(\frac{1}{y_1}) \tau=\frac{\tau}{1+\tau}
\end{eqnarray*}

Predictor-Corrector

\begin{eqnarray*}
k_1=\frac{1}{y_0} \\
k_2=\frac{1}{y_0+k_1 \tau} \\
y_1=y_0+\frac{tau}{2}(1+\frac{1}{1+\tau}) = 1+\frac{2 \tau+\tau^2}{2(1+\tau)}
\end{eqnarray*}

{\bf Problem 2}
\smallskip

Applying one iteration of Euler Explicit to the given system, $\tau=0.1$, gives

\begin{eqnarray*}
y_1=y_0+y'\tau \\
 y_1=\begin{bmatrix}
	1 \\
	0 
	\end{bmatrix}+\tau
	\begin{bmatrix}
	2 & -1  \\[0.2em]
	1 & -1  \\[0.2em]
     \end{bmatrix}
\begin{bmatrix}
	1 \\
	0 
	\end{bmatrix} \\
y_1=\begin{bmatrix}
	0.8 \\
	0.1 
	\end{bmatrix}
\end{eqnarray*}

{\bf Problem  3}
\smallskip

Solving the second order system using Euler Explicit. $y''=-y$, $y(0)=1$, $y'(0)=0$ goes as follows. 

\begin{eqnarray*}
y_1=y_0+y'\tau \\
 y_1=\begin{bmatrix}
	1 \\
	0 
	\end{bmatrix}+\tau
	\begin{bmatrix}
	0 & 1  \\[0.2em]
	-1 & 0  \\[0.2em]
     \end{bmatrix}
\begin{bmatrix}
	1 \\
	0 
	\end{bmatrix} 
\end{eqnarray*} 

We can analyze its stability by looking at the eigenvalues of the matrix representing the differential equation. 

\begin{equation*}
\Big{|}
\begin{bmatrix}
-\lambda & 1  \\
-1 & -\lambda  
\end{bmatrix}
\Big{|}=0
\bigskip
\end{equation*}

From which we see $\lambda= \pm i$. The solution to this system goes as $|1- \lambda \tau |$. 

\begin{equation*}
|1-\lambda \tau|=\sqrt{1+\tau^2}>1
\end{equation*}

Which implies instability. 
\bigskip

{\bf Problem  4}
\smallskip

I'm not entirely sure how to do this problem, but here is my best guess. For the IVP $y'=-y$, $y(0)=1$ we are given $2 \alpha_0+\alpha_1=1$. We can write this system as a matrix

\begin{equation*}
y'_1=\tau  
\begin{bmatrix}
	\alpha_0 & 0 & 0  \\[0.2em]
	0 & \alpha_1 & 0  \\[0.2em]
	0 & 0 &\alpha_ 0  \\[0.2em]
\end{bmatrix}
\begin{bmatrix}
y_0 \\
y_1 \\
y_2 
\end{bmatrix}
\end{equation*}

The eigenvalues of this matrix are $\lambda_1=\alpha_1$, $\lambda_{2,3}=\alpha_0$. If it's still true that $|1-\lambda \tau|<1$, then We must restrict $\alpha$ (since no restriction on $\tau$ is imposed). From this we see that $\alpha$ must be less than 1 but greater than 0. 
\bigskip

{\bf Problem  5}
\smallskip

We apply the classical RK method to the IVP $y'=-\lambda y$, the solution of which is actually $y(t)=e^{-\lambda t}$. After much algebra we can see that using the RK4 method gets us

\begin{equation*}
y_1=y_0-\lambda \tau + \frac{1}{2} \tau^2 \lambda ^2 - \frac{1}{6} \tau^3 \lambda ^3 + \frac{1}{24} \tau^4 \lambda ^4
\end{equation*}

Which is the Taylor series of the actual solution up to the fourth term. In terms of stability the linear term dominates, meaning 

\begin{equation*}
|y_0-\lambda \tau| < 1
\end{equation*}

Since $y_0=1$, $\lambda \tau < 2$ for stability (assuming both are $>$ 0). In other words, $\tau<\frac{2}{\lambda}$. 
\bigskip

{\bf Problem 6}
\smallskip

We apply the explicit predictor-corrector method to the IVP $y'=-\lambda y$, the solution of which is $y(t)=e^{-\lambda t}$. 

\begin{eqnarray*}
y_1=y_0+\tau(\alpha y_0'+\beta(\tau y_0''+\frac{1}{2}\tau^2 y_0''')) \\
y_1=y_0+\tau \alpha y_0'+\tau \beta(\tau y_0''+\frac{1}{2} \tau^2 y_0''') \\
y_1=y_0+\tau \alpha y_0'+\tau^2 \beta y_0'' + \frac{1}{2} \tau^3 y_0'''
\end{eqnarray*}

From matching the Taylor series we can see that $\alpha=-1$, $\beta=\frac{1}{2}$, although we can never eliminate the third term (unless I've made an error.) Again, the linear term governs stability. 

\begin{equation*}
|y_0-\lambda \tau| < 1
\end{equation*}

Since $y_0=1$, $\lambda \tau < 2$ for stability (assuming both are $>$ 0). In other words, $\tau<\frac{2}{\lambda}$, making the stability conditional. 
\bigskip

{\bf Problem 7}
\smallskip

We attempt to approximate the IVP $y'=-y$ with a multistep method, the characteristic equation for which is 

\begin{equation*}
z^2+2\tau z-y_0=0
\end{equation*}

Solving for $z$, we see that 

\begin{eqnarray*}
z=\frac{-2\tau \pm \sqrt{4(\tau^2+1)}}{2} \\
z=-\tau \pm \sqrt{\tau^2+1}
\end{eqnarray*}

In order for this to be stable, $|z|<1$. It is clear that this is impossible because $|-\tau-\sqrt{\tau^2+1}|>1$ for any $\tau$. Therefore this approximation is not stable. 
\bigskip

{\bf Problem 8}

In approximating $y''=-y$, we will use a solution whose characteristic equation is 

\begin{equation*}
z^2+(\tau-2)z+1=0
\end{equation*}

Solving for $z$, we see that 

\begin{eqnarray*}
z=\frac{2-\tau^2 \pm \sqrt{(\tau^2-2)^2-4y_0}}{2}
\end{eqnarray*}

In order for this to be stable, the discriminant $(\tau^2-2)^2-4y_0$ must be negative (make $z$ complex). In order for that to happen, $0 < \tau < 2$. 


\end{document}
